---
title: "Assignment 2: Exponential Random Graph Model"
date: "2025-01-13"
output: 
  pdf_document:    
    latex_engine: xelatex
    keep_tex: true
    number_sections: false
knitr:
  opts_chunk:
    echo: true              
    warning: true           
    message: true           
    fig.align: "center"     
    fig.width: 6            
    fig.height: 4           
editor_options: 
  markdown: 
    wrap: sentence
---

#### By Marco Di Stasio, Alessandra Gandini, Gaudenzia Genoni, Yishak Tadele Nigatu

# Introduction

This study applies ERGM analysis to a directed, valued friendship network of 32 network scientists, derived from the Electronic Information Exchange System (EIES) dataset.
After dichotomizing the network, three models were estimated:

1.  Purely structural model, capturing key network-level features such as reciprocity, degree distribution, and triadic structures;
2.  Dyadic-independent model incorporating gender attributes to assess individual-level effects; and
3.  Combined model integrating both structural and attribute-based effects.

The coefficients of the models are interpreted with a substantial focus on Model 3, followed by goodness-of-fit diagnostics to evaluate how well each model replicates the observed network structure and relational patterns.
The combined model, which includes structural features and gender-related effects, provides the most comprehensive representation of the network.
However, discrepancies in extreme degree values, certain triadic configurations, and long-range connectivity suggest room for refinement to account for additional effects.

## Part One: Load Network and Run Models

```{r import packages, echo=TRUE, message=FALSE}
library(sna)
library(ergm)
library(readxl)
library(knitr)
library(broom)
```

```{r load main data, echo=TRUE}
EIES_T2F_n <- read.csv("data/EIES_T2.csv", row.names=1)
EIES_T2F_n <- EIES_T2F_n > 2 #consider only friends (3) and good friends (4)

EIES_T2F_n<-as.network(EIES_T2F_n, directed=T) #transform into a network object
```

```{r load gender data, fig.align='center'}
Attributes <- read_excel("data/Data_Gender_3.xlsx")
EIES_T2F_n %v% "Gender"<- Attributes$Gender

plot(EIES_T2F_n)
gplot(EIES_T2F_n, gmode="graph", displaylabels=T, vertex.col=EIES_T2F_n %v% "Gender"*2)
```

The study examines acquaintance relationships among 32 network scientists using the EIES dataset at Time 2.
To focus on strong ties, the directed, valued network was dichotomized and transformed into a network: edges with values of 3 (friend) or 4 (close personal friend) were coded as 1, and all others (0, 1, 2) were coded as 0.
Gender attributes (from an external dataset) were then assigned to each node, creating an enriched network object for analysis.

Three exponential random graph models (ERGMs) are estimated.
**Model 1** captures purely structural effects, including edges, mutual ties, geometrically weighted in-degree and out-degree distributions, and two-path dependencies (outgoing and incoming).
**Model 2** examines dyadic independent effects by incorporating individual and dyadic attributes, specifically gender as a nodal covariate for both in-degree and out-degree, as well as gender homophily through nodal matching.
**Model 3** combines structural and attribute-based effects, integrating the terms from both Models 1 and 2 to provide a comprehensive analysis of the network.

### Model 1

```{r train model1, , message=FALSE, fig.show='hide',warning=FALSE, echo=FALSE}
EIES_Model1<-ergm(EIES_T2F_n ~ edges+mutual
                  +gwidegree(decay=.3, fixed=TRUE) +gwodegree(decay=.3, fixed=TRUE)
                  +dgwesp(type="OTP", decay=.5, fixed=TRUE)
                  +dgwesp(type="ITP", decay=.5, fixed=TRUE),
                  control=control.ergm(seed=102, MCMC.runtime.traceplot=TRUE), 
                  verbose=TRUE)
```

```{r model1_summary,echo=FALSE}
summary(EIES_Model1)
```

### Model 2

```{r train_model2, , message=FALSE, fig.show='hide',warning=FALSE, echo=FALSE}
EIES_Model2<-ergm(EIES_T2F_n ~ edges+mutual
                  +nodeicov("Gender")+nodeocov("Gender")+nodematch("Gender"), 
                  control=control.ergm(seed=102, MCMC.runtime.traceplot=TRUE), 
                  verbose=TRUE)
```

```{r model2_summary, echo=FALSE}
summary(EIES_Model2)
```

### Model 3

```{r train model3, message=FALSE, fig.show='hide',warning=FALSE, echo=FALSE}
EIES_Model3<-ergm(EIES_T2F_n ~ edges+mutual
                  +gwidegree(decay=.3, fixed=TRUE) +gwodegree(decay=.3, fixed=TRUE)
                  +dgwesp(type="OTP", decay=.5, fixed=TRUE)
                  +dgwesp(type="ITP", decay=.5, fixed=TRUE)
                  +nodeicov("Gender")+nodeocov("Gender")+nodematch("Gender"), 
                  control=control.ergm(seed=102, MCMC.runtime.traceplot=TRUE), 
                  verbose=TRUE)

```

```{r model3_summary, echo=FALSE}
summary(EIES_Model3)
```

The results are presented in the following table:

| Effect           | Model 1 Estimate (SE) | Model 1 Sign | Model 2 Estimate (SE) | Model 2 Sign | Model 3 Estimate (SE) | Model 3 Sign |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| edges            | -4.825 (0.363)        | \*\*\*       | -2.941 (0.185)        | \*\*\*       | -5.562 (0.446)        | \*\*\*       |
| mutual           | 2.603 (0.344)         | \*\*\*       | 2.486 (0.291)         | \*\*\*       | 2.678 (0.360)         | \*\*\*       |
| gwideg.fixed     | 2.195 (1.135)         | .            | NA                    | NA           | 1.641 (1.164)         |              |
| gwodeg.fixed     | 4.377 (2.792)         |              | NA                    | NA           | 16.286 (7.005)        | \*           |
| gwesp.OTP.fixed  | 1.967 (0.218)         | \*\*\*       | NA                    | NA           | 1.853 (0.223)         | \*\*\*       |
| gwesp.ITP.fixed  | -0.627 (0.110)        | \*\*\*       | NA                    | NA           | -0.637 (0.118)        | \*\*\*       |
| nodeicov.Gender  | NA                    | NA           | 0.053 (0.217)         |              | 0.005 (0.189)         |              |
| nodeocov.Gender  | NA                    | NA           | 1.274 (0.210)         | \*\*\*       | 1.236 (0.309)         | \*\*\*       |
| nodematch.Gender | NA                    | NA           | -0.027 (0.162)        |              | -0.071 (0.161)        |              |

## Part Two: Discussion of ERGM Results

In Model 3, Five effects are significant (edges, mutual, gwodeg.fixed, gwesp.OTP.fixed, and gwesp.ITP.fixed), while the remaining three (gwideg.fixed, nodeicov.Gender, and nodematch.Gender) are not.

Below, we discuss **Model 3** in detail and highlight notable differences from **Models 1** (structural only) and **2** (attributes only).

### Model 3 Descriptions

1.  **Edges**
    -   **Estimate (SE)**: **-5.562 (0.446)**, **p \< 0.001**
    -   **Interpretation**: The strong negative coefficient indicates a low baseline propensity for any given pair to be "friends". In a friendship context, this is typical because only a minority of all possible ties are strong friendships. This could also be a result of the sparsity of the network.
2.  **Mutual**
    -   **Estimate (SE)**: **2.678 (0.360)**, **p \< 0.001**
    -   **Interpretation**: This significant positive effect suggests a strong tendency for reciprocity. In other words, if Person A calls Person B a friend, Person B is much more likely to call Person A a friend in return. This effect remains comparably large and significant across all models (it is also highly significant in **Models 1** and **2**).
3.  **gwidegree**
    -   **Estimate (SE)**: **1.641 (1.164)**, **not significant**
    -   **Interpretation**: This parameter captures "popularity" (i.e., whether certain nodes consistently receive many friendship nominations). Although it was marginally significant in **Model 1** (p ≈ 0.05) and still positive here, it is not statistically significant in **Model 3**. After accounting for other structural and gender effects, there is no strong residual tendency for some individuals to be universally popular as "friends."
4.  **gwodegree**
    -   **Estimate (SE)**: **16.286 (7.005)**, **p \< 0.05**
    -   **Interpretation**: This effect captures "activity" (i.e., sending out multiple friendship nominations). While this effect was positive but not statistically significant in **Model 1**, it becomes both larger and significant in **Model 3**, implying that once we control for gender and other structural factors, some individuals indeed nominate friends at a higher‐than‐expected rate. Substantively, this suggests a group of "active connectors" who name multiple close colleagues as friends.
5.  **dgwesp.OTP**
    -   **Estimate (SE)**: **1.853 (0.223)**, **p \< 0.001**
    -   **Interpretation**: A positive and significant OTP parameter indicates that triadic closure is especially likely when an individual sends ties to two different people (i.e., "friends of my out‐friends tend to be my friends too"). This finding aligns with typical friendship processes where one's out‐contacts may be more likely to know each other, and it is very similar in magnitude to **Model 1**, implying that adding gender attributes does not substantially alter this triadic closure dynamic.
6.  **dgwesp.ITP**
    -   **Estimate (SE)**: **-0.637 (0.118)**, **p \< 0.001**
    -   **Interpretation**: In contrast, the negative and significant ITP parameter implies that if two people both receive ties from the same individual, they are *less* likely than random to be tied themselves (once we account for all other effects). In friendship terms, this suggests that people do not necessarily become friends merely because they share a common "admirer." This pattern remains consistent from **Model 1** through **Model 3**.
7.  **nodeicov("Gender") - male**
    -   **Estimate (SE)**: **0.005 (0.189)**, **not significant**
    -   **Interpretation**: This coefficient tests whether being **male(coded as 0)** makes one more likely to *receive* friend nominations. It is essentially zero and non‐significant, indicating that once we control for other network tendencies, male actors do not appear to receive significantly more (or fewer) friendship ties than female actors.
8.  **nodeocov("Gender"=1) - female**
    -   **Estimate (SE)**: **1.236 (0.309)**, **p \< 0.001**
    -   **Interpretation**: This significant positive coefficient indicates that **females(coded as 1)** are more likely to send friendship ties than males. Notably, it was also significant in **Model 2** at a similar magnitude (1.274). Thus, even after accounting for structural processes like reciprocity and triadic closure, women in this network remain more active in naming friends than men.
9.  **nodematch("Gender"=0)**
    -   **Estimate (SE)**: **-0.071 (0.161)**, **not significant**
    -   **Interpretation**: This nonsignificant coefficient suggests that there is no strong homophily or heterophily by gender for close friendships---i.e., men do not show a strong tendency to befriend men, nor women to befriend women, relative to cross‐gender friendships, once other factors are considered. This is consistent across **Models 2** and **3**.

### Notable Differences from Models 1 and 2 to Model 3

-   **gwodegree**: Became statistically significant in **Model 3** but was not in **Model 1**(β=4.372,p=0.117), highlighting that individuals who send many ties ("outgoing activity") do not significantly influence the overall network structure at this stage. However, when gender attributes are introduced in Model 3, individuals who actively initiate multiple friendships (high out-degree) appear to play an important role in shaping the network, especially when considering the sender's gender. This could reflect that certain individuals are more proactive in initiating friendships, and their role becomes more evident when both structural and gender-related effects are accounted for.
-   **Edges**: Became more strongly negative in **Model 3** (--5.562) than in **Models 1** or **2**, a typical pattern when multiple positive interaction processes (e.g., reciprocity, triadic closure) are added.
-   **gwidegree**: Marginally significant in **Model 1** (p ≈ 0.05, β=2.195), then not significant when attributes were included in **Model 3**. This suggests that individuals who receive many ties (are "popular") have a slightly increased likelihood of forming additional friendships. When gender attributes are introduced in Model 3, the previously observed in-degree popularity effect is partly explained by other structural or gender-related factors. This suggests that "popularity" in terms of receiving ties is less crucial in explaining tie formation when accounting for sender gender and other structural dynamics. Friendship in this network may not be strongly driven by central hubs but instead distributed across more reciprocal or cohesive relationships.

Generally, **Model 3** demonstrates that both structural forces (reciprocity, triadic closure, activity patterns) and gender attributes shape who becomes friends in this network, though some anticipated effects (like gender‐based popularity or homophily) do not appear to be present.

## Part 3: Goodness of Fit

When performing an ERGM analysis, it is of crucial importance to examine the goodness of fit, as it evaluates how well the model captures the observed network's structure and relational patterns.
Here, goodness-of-fit (GOF) diagnostics were performed by setting a random seed to ensure reproducibility across runs.
Subsequently, 2000 networks were simulated for comparison with the observed network, using the same seed for consistency.
The GOF was calculated for the three models, comparing observed network statistics with those simulated from the model.
The diagnostics include: - *model*: it evaluates how well the model itself fits the data.
- *idegree and odegree*: it assesses the in-degree and out-degree distributions.
- *distance*: it examines the geodesic distance distribution (shortest path lengths between nodes).
- *triadcensus*: it evaluates the distribution of triadic configurations in the network.
These additional metrics (degree distributions, distance, and triadic configurations) are used to complement the evaluation of the model fit.
While the model statistic assesses overall fit, the additional metrics help pinpoint specific network features the model may under- or overestimate.

```{r goodness of fit declaration}
set.seed(565)
gof.choices<-control.gof.ergm(nsim=2000, seed = 565)
```

```{r model1 gof}
#Model 1
EIES_Model1sim2000<-gof(EIES_Model1, 
                        GOF=~model+idegree+odegree+distance+triadcensus, control=gof.choices)
EIES_Model1sim2000$summary.model
```

Overall, the goodness-of-fit diagnostics for Model 1 indicate that the observed network statistics align closely with the simulated values across all metrics.
This suggests that Model 1 provides an adequate fit for capturing the structural properties of the observed network.

```{r model2 gof}
#Model 2
EIES_Model2sim2000<-gof(EIES_Model2, 
                        GOF=~model+idegree+odegree+distance+triadcensus, control=gof.choices)
EIES_Model2sim2000$summary.model
```

The goodness-of-fit diagnostics for Model 2 demonstrate that the observed network statistics are well-replicated by the simulated networks, too.

```{r model3 gof}
#Model 3
EIES_Model3sim2000<-gof(EIES_Model3, 
                        GOF=~model+idegree+odegree+distance+triadcensus, control=gof.choices)
EIES_Model3sim2000$summary.model
```

In the goodness-of-fit diagnostic, each histogram compares an observed network statistic (marked by a vertical red line) to the distribution of that same statistic in a large set of networks simulated from the fitted model.
If the observed value lies near the center of the distribution, then the model is successfully reproducing that feature of the network.
Conversely, if the observed value is off to one side or falls outside the simulated range, it indicates that the model may not be capturing that particular aspect of the network very well.

The goodness-of-fit diagnostics for Model 3 indicate a strong alignment between the observed network statistics and the simulated values, as reflected by high MC p-values across all metrics.
Structural effects, such as edges, mutual ties, and measures of triadic closure (gwesp.OTP.fixed.0.5 and gwesp.ITP.fixed.0.5), are well-captured by the model, showing minimal discrepancies between observed and simulated values.
Additionally, gender-related covariates, including nodeicov.Gender, nodeocov.Gender, and nodematch.Gender, are also accurately represented, suggesting that the model effectively incorporates both structural and attribute-based dynamics.
This demonstrates that Model 3 provides a robust and comprehensive fit to the observed friendship network.

These findings are shown in the histograms below.

```{r histograms, fig.width=30, fig.height=15, echo=FALSE}
# Set up the plotting area to have 3 rows and 3 columns
par(mfrow = c(3, 3))

# Histogram of edges
hist(EIES_Model3sim2000$sim.model[,1] + 0.01, nclass = 20, main = "Histogram of edges", probability = TRUE, xlab = NA)
abline(v = EIES_Model3sim2000$summary.model[1, 1], col = "red", lwd = 3)
abline(v = EIES_Model3sim2000$summary.model[1, 3], col = "blue", lwd = 3, lty = 2)

# Histogram of mutual
hist(EIES_Model3sim2000$sim.model[,2] + 0.01, nclass = 25, main = "Histogram of mutual", probability = TRUE, xlab = NA)
abline(v = EIES_Model3sim2000$summary.model[2, 1], col = "red", lwd = 3)
abline(v = EIES_Model3sim2000$summary.model[2, 3], col = "blue", lwd = 3, lty = 2)

# Histogram of gwideg
hist(EIES_Model3sim2000$sim.model[,3] + 0.01, nclass = 30, main = "Histogram of gwideg", probability = TRUE, xlab = NA)
abline(v = EIES_Model3sim2000$summary.model[3, 1], col = "red", lwd = 3)
abline(v = EIES_Model3sim2000$summary.model[3, 3], col = "blue", lwd = 3, lty = 2)

# Histogram of gwodeg
hist(EIES_Model3sim2000$sim.model[,4] + 0.01, nclass = 30, main = "Histogram of gwodeg", probability = TRUE, xlab = NA)
abline(v = EIES_Model3sim2000$summary.model[4, 1], col = "red", lwd = 3)
abline(v = EIES_Model3sim2000$summary.model[4, 3], col = "blue", lwd = 3, lty = 2)

# Histogram of gwesp.OTP
hist(EIES_Model3sim2000$sim.model[,5] + 0.01, nclass = 30, main = "Histogram of gwesp.OTP", probability = TRUE, xlab = NA)
abline(v = EIES_Model3sim2000$summary.model[5, 1], col = "red", lwd = 3)
abline(v = EIES_Model3sim2000$summary.model[5, 3], col = "blue", lwd = 3, lty = 2)

# Histogram of gwesp.ITP
hist(EIES_Model3sim2000$sim.model[,6] + 0.01, nclass = 30, main = "Histogram of gwesp.ITP", probability = TRUE, xlab = NA)
abline(v = EIES_Model3sim2000$summary.model[6, 1], col = "red", lwd = 3)
abline(v = EIES_Model3sim2000$summary.model[6, 3], col = "blue", lwd = 3, lty = 2)

# Histogram of nodeicov.Gender
hist(EIES_Model3sim2000$sim.model[,7] + 0.01, nclass = 30, main = "Histogram of nodeicov.Gender", probability = TRUE, xlab = NA)
abline(v = EIES_Model3sim2000$summary.model[7, 1], col = "red", lwd = 3)
abline(v = EIES_Model3sim2000$summary.model[7, 3], col = "blue", lwd = 3, lty = 2)

# Histogram of nodeocov.Gender
hist(EIES_Model3sim2000$sim.model[,8] + 0.01, nclass = 30, main = "Histogram of nodeocov.Gender", probability = TRUE, xlab = NA)
abline(v = EIES_Model3sim2000$summary.model[8, 1], col = "red", lwd = 3)
abline(v = EIES_Model3sim2000$summary.model[8, 3], col = "blue", lwd = 3, lty = 2)

# Histogram of nodematch.Gender
hist(EIES_Model3sim2000$sim.model[,9] + 0.01, nclass = 30, main = "Histogram of nodematch.Gender", probability = TRUE, xlab = NA)
abline(v = EIES_Model3sim2000$summary.model[9, 1], col = "red", lwd = 3)
abline(v = EIES_Model3sim2000$summary.model[9, 3], col = "blue", lwd = 3, lty = 2)

```

The *trace plots* showed below are visual diagnostics used to assess the convergence of the Markov Chain Monte Carlo (MCMC) algorithm in Model 3.
In a well-behaved chain, we expect the sampled values to fluctuate around some stable mean (the chain\'s stationary distribution) without systematic trends upward or downward.
The presence of major trends, drifts, or extremely prolonged excursions could indicate that the sampler has not \"mixed\" well or that convergence to a stable distribution has not yet occurred.
By contrast, if the samples appear roughly like a tight, random cloud of points---oscillating within a limited range but neither wandering away nor getting stuck at extremes---this generally suggests good mixing and supports the notion that the MCMC process has reached stationarity.

In these plots, each statistic (e.g., the number of edges or mutual ties in a simulated network) appears to remain within a relatively consistent band of values across the 2,000 iterations.
For instance, the trace for **edges** hovers between roughly 160 and 240, showing no noticeable increasing or decreasing trend over time.
Similarly, **mutual** shifts between about 40 and 80, and the plots for **gwideg** and **gwodeg** fluctuate around stable mid-40s and mid-42s, respectively, without drifting away.
The **gwesp** parameters (OTP and ITP) also move within their respective ranges, and the gender covariates (both nodeicov, nodeocov, and nodematch for Gender) do not show signs of a systematic trend or long-term drift.

Overall, the absence of any major trends, plateaus, or extremely abrupt jumps indicates that the MCMC procedure is performing reliably for each term in the model.
This reliability underpins the validity of the fitted ERGM and supports the interpretation that Model 3 has converged, providing stable parameter estimates that capture the observed network structure and attribute effects.

```{r trace plots, fig.width=40, fig.height=30, echo=FALSE}

par(mfrow = c(3, 3))

plot(EIES_Model3sim2000$sim.model[,1], type="l", main = paste("Trace plot for edges"), ylab="", xlab="")

plot(EIES_Model3sim2000$sim.model[,2], type="l", main = paste("Trace plot for mutual"), ylab="", xlab="")

plot(EIES_Model3sim2000$sim.model[,3], type="l", main = paste("Trace plot for gwideg"), ylab="", xlab="")

plot(EIES_Model3sim2000$sim.model[,4], type="l", main = paste("Trace plot for gwodeg"), ylab="", xlab="")

plot(EIES_Model3sim2000$sim.model[,5], type="l", main = paste("Trace plot for gwesp.OTP"), ylab="", xlab="")

plot(EIES_Model3sim2000$sim.model[,6], type="l", main = paste("Trace plot for gwesp.ITP"), ylab="", xlab="")

plot(EIES_Model3sim2000$sim.model[,7], type="l", main = paste("Trace plot for nodeicov.Gender "), ylab="", xlab="")

plot(EIES_Model3sim2000$sim.model[,8], type="l", main = paste("Trace plot for nodeocov.Gender "), ylab="", xlab="")

plot(EIES_Model3sim2000$sim.model[,9], type="l", main = paste("Trace plot for nodematch.Gender "), ylab="", xlab="")
```

Finally, the evaluation of the model fit is completed with the analysis of additional metrics, along with a re-presentation of the GOF for model statistics (which was already discussed above).

```{r goodness of fit diagnostic, fig.align='center', echo=FALSE}
par(mfrow = c(2,3))
plot(EIES_Model3sim2000)
```

The goodness-of-fit diagnostics for the additional metrics in Model 3 reveal varying levels of alignment between the observed and simulated network statistics, as detailed below for each of them.

#### Boxplots of Key Statistics (Edges, Mutual, gwodeg, gwesp, nodecov.Gender, etc.)

The fitted model reproduces the fundamental network features very well, evidenced by high MC p‐values for edges (p=1.000), mutual ties (p=0.979), and other key statistics (gwideg, gwodeg, gwesp, and nodematch by Gender).
For instance, the observed number of edges (204) sits comfortably within the simulated range (159 to 250), and the same holds for mutual ties (observed=60, mean=59.75).
These results indicate that the model captures both basic connectivity (edges) and more nuanced structural tendencies (reciprocity, geometrically weighted terms, gender effects).

#### Out‐Degree and In‐Degree Distributions

The second and third plots show the *distribution* of out‐degree (number of ties sent) and in‐degree (number of ties received) across all nodes.
The outdegree plot focuses on how many friends each person *nominates*, whereas the indegree plot shows how many friend nominations each person *receives*.

Overall, the model does a good job fitting most in‐degree categories, as shown by high p‐values for degrees such as 6 (p=0.926), 7 (p=1.000), and 10 (p=1.000).
However, certain counts depart from the observed data.
For example, the frequency of nodes with 1 incoming tie (observed=3) is underestimated (mean=1.1410; p=0.207), whereas the number of nodes with 5 incoming ties (observed=1) is overestimated (mean=3.8815; p=0.169).
The biggest mismatch occurs at the extreme end (idegree24): the model never predicts this very high in‐degree (mean=0.0000; p=0.000), suggesting it struggles with nodes that receive a very large number of ties.

The out‐degree distribution also aligns well with the observed data in most categories, reflected in high p‐values for many degrees (e.g., odegree4 p=1.000, odegree6 p=1.000).
Nonetheless, the model underestimates some degrees (e.g., odegree3 observed=2 vs. mean=5.9685, p=0.068) and slightly overestimates others (e.g., odegree7 p=0.079, odegree9 p=0.069).
A notable discrepancy occurs at odegree20 (observed=1 vs. mean=0.028, p=0.056), indicating the model struggles to capture the presence of a rare "super‐connector."

### Triad Census

The triad census (fourth plot) displays the proportion of all triads (three‐node subgraphs) of each type---e.g., \"003\" means no ties among the three nodes, \"021D\" means a particular pattern of two directed edges, \"111U\" a fully mutual triad, and so on.
The black line (or diamonds) often represents the observed proportions of each triad type; the boxes/whiskers indicate the simulated range.

Most triad types appear in realistic proportions (e.g., 003 has p=0.693, 102 p=0.874), which suggests the model handles common triad structures---no ties, single arcs, and so forth---adequately.
However, certain directed triads are misestimated.
For instance, 021U (p=0.000) and 021C (p=0.001) deviate markedly from the observed data(can be seen from the following picture), and fully reciprocal triads (111D, 111U) also show some discrepancies (p=0.040 and 0.218, respectively).
Although overall triadic structure is captured, those particular directed subgraphs remain more challenging for the model.

```{r, fig.width=8, fig.height=6, echo=FALSE, fig.align='center'}
temp <- c(5, 6)
par(mfrow = c(2, 1))

for (k in temp) {
  hist(
    EIES_Model3sim2000$sim.triadcensus[, k],
    main = paste("Poor Estimates for ", colnames(EIES_Model3sim2000$sim.triadcensus)[k])
  )
  abline(
    v   = EIES_Model3sim2000$obs.triadcensus[k],
    col = "blue",
    lwd = 3,
    lty = 2
  )
}

```

### Geodesic Distance Distribution

In the final plot, each dyad\'s minimum geodesic distance (the shortest path length between two people) is tabulated and grouped, then compared across observed vs. simulated networks.
Distances of 1 mean direct ties; distance 2 means you can reach one person via a single intermediary, and so forth.

Distances of 1, 5, 6, 7, 8, and etc match closely between observation and simulation, as indicated by p‐values near 1.0.
Distance 2 (observed=443 vs. mean=483.53, p=0.310) and distance 3 (observed=262 vs. mean=232.94, p=0.417) also fall within acceptable ranges.
The largest gap arises at distance 4 (observed=82, mean=40.29, p=0.126), suggesting the model underestimates the number of paths requiring four steps.
Even so, the overall shape of the distance distribution is well captured, meaning the model generally reflects how "far apart" nodes are in this network.

```{r EIES_Model3sim2000 summary}
EIES_Model3sim2000
```

In conclusion, the model is satisfactory in capturing the overall structure and many relational patterns of the network, particularly those related to reciprocity, clustering, and the influence of gender attributes.
However, its limitations in handling extreme degree values, specific triadic structures, and longer paths suggest that additional effects may need to be included to improve the model's fit.
These could include effects related to preferential attachment, node-specific roles, or alternative triadic configurations, which might better capture the underlying processes driving tie formation in the network.

### Bibliography

Borgatti, S. P., Everett, M. G., Johnson, J. C., & Agneessens, F.
(2022).
Analyzing social networks using R.
SAGE Publications.
